---
title: "Supplemental Script from J. Proteome Res. article"
author: "Kai-Ting Fan, adapted by Aaron Rendahl"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{J Proteome Res Supplemental Script}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
 knitr::opts_chunk$set(eval=FALSE)
```

This vignette contains the code the supplemental script from the J. Proteome Res. article (DOI:10.1021/acs.jproteome.5b00772). After the ProteinTurnover package is installed, the location of the script can be found using the `system.file` command.

```{r, eval=FALSE}
system.file(package="ProteinTurnover", "extscripts", "JProteomeRes_Supplemental.R")
```

## SECTION 1: Install R packages
Install package "ProteinTurnover" and other required packages.
R-3.2.0 or higher is required.

To access mzXML files, the package "xcms" from Bioconductor is also required. For installation:
```{r, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("xcms")  
```

The ProteinTurnover package is available on GitHub, https://github.com/HegemanLab/ProteinTurnover.
```{r, eval=FALSE}
devtools::install_github(HegemanLab/ProteinTurnover)
```

## SECTION 2: Prepare data files

* All MS data files were converted from .RAW to .mzXML files.
* Peptide identification file was exported from Scaffold spectrum report and saved as .csv file.
* This .csv file needs to contain peptide retention time in column "Spectrum name" following this format "Elution: xx min".
* Please also refer Supporting Information Table S-3 for the required column headers in the .csv file.
* All data files should be placed in the same data directory.

## SECTION 3: Run ProteinTurnover
Load library.
```{r}
library(ProteinTurnover)
```

Set up list to hold parameters for this run.
```{r}
p <- list()
```

Set the number of cores to use. 
On Windows OS, this should be 1; on Linux, it can be set as 8.
```{r}
p$num.cores <- 1
```

Set directories and scaffold file name:

* Set the data file directory.
* Set the directory that result files should be stored in.
* Set scaffold file, which should be store in the data directory.
* Note: For Windows OS users, replace all the \ with \\ from the copy of directory path.

This is an example data directory:
```{r}
p$dir.data <- "C:\\Users\\USERNAME\\Documents\\mzs"  
```

This is an example result directory:
```{r}
p$dir.results <- "C:\\Users\\USERNAME\\Documents\\Turnover"  
```

This is an example peptide file:
```{r}
p$scaffoldfile <- "soluble-FDR.csv"   
```

Set times and corresponding file names:

* Set times as the labeling time points.
* set names of mzXML files that correspond to the above time points.
* set names of rds (R data files) that correspond to the above times.
* In this example, mzXML files in the data directory are named as hr_0.mzXML, hr_8.mzXML, etc. by users.
* Note: R is case sensitive, including the file extension name.

```{r}
p$times <- c(0,4,8,16,24,32,40,48)
p$rawfiles <- paste0("hr_", p$times, ".mzXML") 
p$rdsfiles <- paste0("hr_", p$times, ".rds")
```

Set parameters for use in the fit.

* The confidence level is based on the peptide identification confidence value.
* m/z windows set as plus or minus 0.005
* Retention time window set as plus or minus 30 sec
* Add extra 5 isotopic channels
```{r}
p$isotope <- "N"
p$confidence.level <- 80  
p$time.zero <- 0
p$time.unit <- "Hour" 
p$mz.tol <- 0.005
p$rt.tol <- 30
p$extra.channels <- 5
```

Set parameter to get the relative abundance from EIC (relAbForTimes).
Available options are ("lm", "rlm", "lqs", "rq", "sum","log").
```{r}
p$regression.model <- "lm"   # "lm" as linear regression
```

Set up parameters for fitting. Options are "incorporation" for stable isotope label incorporation and "dilution" for label dilution
```{r}
p$isotope.method <- "incorporation"
```

Set up betabinomial options for the "M" parameter in Eq.(5).
Available options are ("none", "one", "many"). 
If users choose "none", then the binomial model will be used
```{r}
p$M.model <- "one"
```

Set the alpha parameters, Eq.(6).
For incorporation, available options are ("many", "k", "kplateau", or "log2k").
For dilution, available options are ("one", "many").
"many" always means the algorithm will fit a different value of that parameter for each time point.
```{r}
p$alpha.model <- "log2k"     #make log2 value of turnover rate k
```

This is passed to the "nab" parameter of pepcurve.
```{r}
p$plim <- NA
```

### Create results directory
This will give warning if already exists.
```{r}
dir.create(p$dir.results)
```

### Convert raw mzXML files to Rdata files
This only need be done once.
This speeds up the steps below; it can be skipped if the raw files are used in addEICs.
```{r}
rawToRDS(p$dir.data, p$rawfiles, p$rdsfiles)
```

### Read in all the data
Read the scaffold file.
```{r}
dat <- readScaffoldFile(p$scaffoldfile, times=p$times, dir=p$dir.data)
```

Prepare the data structure for each sequence.
```{r}
seqs <- makeSequences(dat,
                      isotope=p$isotope,
                      confidence=p$confidence.level,
                      time.zero=p$time.zero,
                      time.unit=p$time.unit,                      
                      mz.tol=p$mz.tol,
                      rt.tol=p$rt.tol,
                      extra.channels=p$extra.channels)
```

Get the EIC data for each sequence.
```{r}
seqs <- addEICs(seqs, files=p$rdsfiles, type="rds", times=p$times, dir=p$dir.data)
```

Save what we've done so far.
```{r}
save(dat, seqs, p, file=file.path(p$dir.results, "eicdata.Rdata"))
```

### Get fits for each sequence
Reload what we've done so far.
Skip these two command lines if users just run the above code
```{r}
dir.results <- "C:\\Users\\USERNAME\\Documents\\Turnover"  
load(file.path(dir.results, "eicdata.Rdata"))       
```

Set up multiple cores.
```{r}
if(!is.null(p$num.cores)) { options(mc.cores=p$num.cores) } 
```

Get the distribution abundance proportion and fit for each sequence.
mclapply.progress is just like lapply except works in parallel and shows progress.
makeRelAb is a wrapper for relAbForTimes, to handle any error.
makeFit is a wrapper for pepfit, to handle any error.

The algorithm chooses parameters that make the smallest difference between the observed proportion for each isotopic channel to the predicted proportion.
In makeFit, if users prefer using relative abundance instead of proportion (the current default setting) for any reason, they can add "useRelAb=TRUE" as a new parameter.
Across time points, the proportionality constant differs, so using relative abundance gives more weight to the time points with higher total values of relative abundance.
Using proportions will treat each time point equally.

```{r}
seqs <- mclapply.progress(seqs, function(seq) {
  relAb <- makeRelAb(seq, regression.model=p$regression.model)
  fit <- makeFit(seq=seq, relAb=relAb, M=p$M.model, alpha=p$alpha.model, se=TRUE, isotope.method=p$isotope.method)
  list(seq=seq, relAb=relAb, fit=fit)    
})
```

Save results so far.
```{r}
save(dat, seqs, p, file=file.path(p$dir.results, "fit-log2k.Rdata"))
```


### Get parameters and save to a file
Load results.
Skip these two command lines if users just run the above code.
```{r}
dir.results <- "C:\\Users\\USERNAME\\Documents\\Turnover"  
load(file.path(dir.results, "fit-log2k.Rdata"))     
```

Get the parameter fits for those succeeded and write to file.
```{r}
fits <- getPars(seqs)
write.csv(fits, file=file.path(p$dir.results, "fits-log2k.csv"), na="")
```

### Make HTML output for each seq

Load what we've done so far.
Skip these two command lines if users just run the above code.
```{r}
dir.results <- "C:\\Users\\USERNAME\\Documents\\Turnover"  
load(file.path(dir.results, "fit-log2k.Rdata"))     
```

Set up multiple cores.
```{r}
if(!is.null(p$num.cores)) { options(mc.cores=p$num.cores) } 
```

Make the output.
```{r}
out <- makeSequenceHTML(seqs, dir=p$dir.results, file="results-log2k")
```


## SECTION 4: Explore each peptide's fit as desired
Load saved result.
```{r}
dir.results <- "C:\\Users\\USERNAME\\Documents\\Turnover"  
load(file.path(dir.results, "fit-log2k.Rdata"))
```

Pick a fit to explore.
Put the index number of whichever peptide you want to explore, eg. No.2 peptide in the output file.
```{r}
s <- seqs[[2]]
```

Get usual output for the fit.
```{r}
s$fit
score.visual(s$fit)
```

Make desired pictures for this No.2 peptide:

Make regression plots:
```{r}
regressionPlot(s$relAb)
```

Make eic plots:
```{r}
plot(s$relAb)
```

Make MLE plots:
```{r}
plot(s$fit)
```

Make the plot of non-linear regression curve:
```{r}
plot(s$curve)
```

## SECTION 5: Use alpha model as "many" instead

Load saved EIC data.
```{r}
dir.results <- "C:\\Users\\USERNAME\\Documents\\Turnover"  
load(file.path(dir.results, "eicdata.Rdata"))
```

Set alpha model differently.
```{r}
p$alpha.model <- "many"
```

Set up multiple cores.
```{r}
if(!is.null(p$num.cores)) { options(mc.cores=p$num.cores) } 
```

Get the distribution abundance proportion, fit, and curve for each sequence.
```{r}
seqs <- mclapply.progress(seqs, function(seq) {
  relAb <- makeRelAb(seq, regression.model=p$regression.model)
  fit <- makeFit(seq=seq, relAb=relAb, M=p$M.model, alpha=p$alpha.model, se=TRUE, isotope.method=p$isotope.method)
  list(seq=seq, relAb=relAb, fit=fit)    
  curve <- makeCurve(fit, plim=p$plim, isotope.method=p$isotope.method)
  list(seq=seq, relAb=relAb, fit=fit, curve=curve)
})
```

Save results so far.
```{r}
save(dat, seqs, p, file=file.path(p$dir.results, "fit-many.Rdata"))
```

Get the parameter fits for those succeeded and write to file.
```{r}
fits <- getPars(seqs)
write.csv(fits, file=file.path(p$dir.results, "fits-many.csv"), na="")
```

Output to HTML:
```{r}
out <- makeSequenceHTML(seqs, dir=p$dir.results, file="results-many")
```
